---
title: "DBMS Practicum 2"
author: "Sakthi Kripa Selvan (selvan.sa@northeastern.edu), Srikanth Kadaba Bhogananda (kadababhogananda.s@northeastern.edu)"
date: "05/03/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
---

# Part 2

#### Loading the required library

```{r libraries, message=FALSE}
library(RMySQL)
library(RSQLite)
library(DBI)
library(sqldf)
library(ggplot2)
db_user <- 'admin'
db_password <- 'sakthi123'
db_name <- 'practicum2db'
db_host <- 'practicum2db.cozotp8befyg.us-east-1.rds.amazonaws.com'
db_port <- 3306
```


#### Establishing connection with the MySQL database.

```{r}
mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                  dbname = db_name, host = db_host, port = db_port)
```

```{r}
detach("package:RMySQL")
library(sqldf)
```

Connecting with the SQLite database to connect with the db created in Part1

```{r}
fpath = "./"
file = "pubmed.db"
con <- dbConnect(RSQLite::SQLite(), paste0(fpath, file))
```

#### Preparation of data for authorFact table

Using the tables in Part 1 to create a authorFacts table as requested. This table has data in the level of auid (i.e) auid is the primary key for the table. It also has the author name, the number of publications by the author and the average publications per year by the author. 

```{r}
sqlCmd = "SELECT au.auid, 
                 (au.FirstName ||' '|| au.LastName ||' '|| Initials) as authorName,
                 COUNT(*) as TotalPublications,
                 ROUND(((COUNT(*)*1.0)/(SELECT COUNT(DISTINCT PublishYear) as CalcField 
                       FROM Journals)),2) as AveragePublications
          FROM Authors au
          JOIN authorArticleMap map on map.auid = au.auid
          JOIN Articles ar on map.PMID = ar.PMID
          GROUP BY au.auid, (au.FirstName ||' '|| au.LastName ||' '|| Initials)"

authorFacts_df = dbGetQuery(con, sqlCmd)
head(authorFacts_df)
```

#### Fact table creation in MySQL

For this part of the practicum, a single fact table is created with all the attributes asked for, eliminating the need for dimension table. 

```{sql connection=mydb}
DROP TABLE IF EXISTS authorFacts;
```

```{sql connection=mydb}
CREATE TABLE authorFacts(
auid INT NOT NULL,
authorName VARCHAR(255),
TotalPublications INT,
AveragePublications FLOAT,
PRIMARY KEY (auid)
)
```

#### Loading data into the table

```{r}
# dbWriteTable is used with append to insert the data from data frame into table.
dbWriteTable(mydb, 'authorFacts', authorFacts_df, row.names=F,append=T, overwrite=F)
```

#### Display the first few rows

```{sql connection=mydb}
SELECT * FROM authorFacts
       LIMIT 5;
```

# Part 3

#### Top 10 Authors by number of publications
The following query into the fact table fetches the top 10 authors by the number of publications. 
```{r}
topAuthorsdf <- dbGetQuery(mydb, statement = paste("SELECT authorName, TotalPublications
                                                FROM authorFacts
                                                ORDER BY TotalPublications DESC, authorName
                                                LIMIT 10"))
topAuthorsdf
```
No updates are required to the fact table to accommodate the above query.

#### Seasonal Patterns in number of publications

To observe the seasonal patterns in the number of publications, we need to include information on publish year and month into the authorFacts table. After this update, auid, publishyear and publishmonth together are the primary keys for this table. Below is the SQL query used to get the columns required. 

```{r}
sqlCmd = "SELECT au.auid, 
                 (au.FirstName ||' '|| au.LastName ||' '|| Initials) as authorName,
                  CAST (PublishYear AS CHAR) PublishYear,
                 PublishMonth,
                 COUNT(*) as TotalPublications,
                 ROUND(((COUNT(*)*1.0)/(SELECT COUNT(DISTINCT PublishYear) as CalcField 
                       FROM Journals)),2) as AveragePublications
          FROM Authors au
          JOIN authorArticleMap map on map.auid = au.auid
          JOIN Articles ar on map.PMID = ar.PMID
          JOIN Journals j on j.ISSN = ar.ISSN
          GROUP BY au.auid, (au.FirstName ||' '|| au.LastName ||' '|| Initials), PublishYear,
                   PublishMonth"

authorFactsUpd_df = dbGetQuery(con, sqlCmd)
head(authorFactsUpd_df)
```

#### Creating the authorFactsUpd table 

```{sql connection=mydb}
DROP TABLE IF EXISTS authorFactsUpd;
```

```{sql connection=mydb}
CREATE TABLE authorFactsUpd(
auid INT NOT NULL,
authorName VARCHAR(255),
PublishYear VARCHAR(255) NOT NULL,
PublishMonth INT NOT NULL,
TotalPublications INT,
AveragePublications FLOAT,
PRIMARY KEY (auid, PublishYear, PublishMonth)
)
```

#### Loading data into the table

```{r}
# dbWriteTable is used with append to insert the data from data frame into table.
dbWriteTable(mydb, 'authorFactsUpd', authorFactsUpd_df, row.names=F,append=T, overwrite=F)
```

#### Display the first few rows of the table

```{sql connection=mydb}
SELECT * FROM authorFactsUpd
       LIMIT 5;
```

#### Quarterly trends

Using the month column to get the quarter in which the article was published. 

```{r}
quarterdf <- dbGetQuery(mydb, statement = paste("WITH CTE AS 
            (SELECT *, CASE WHEN PublishMonth <= 3 THEN '1' 
            WHEN PublishMonth >= 4 AND PublishMonth <= 6 THEN '2' 
            WHEN PublishMonth >= 7 AND PublishMonth <= 9 THEN '3' 
            WHEN PublishMonth >= 10 THEN '4' 
            END AS Quarter
            FROM authorFactsUpd
            ORDER BY 1)
            SELECT DISTINCT Quarter, SUM(TotalPublications) OVER(PARTITION BY Quarter) as        
            PublicationsPerQuarter FROM CTE
            ORDER BY 1
            "))
quarterdf
```

```{r}
ggplot(quarterdf, aes(x=Quarter, y=PublicationsPerQuarter, group = 1)) +
  geom_line(color="#69b3a2", size=1) +
  ggtitle("Number of journals published per quarter") +
geom_text(aes(label=PublicationsPerQuarter), vjust=-.5)
```

The highest number of journals published is during the 4th quarter. Publications seem to reduce during the second and third quarter and increase in the final quarter.

#### Yearly trends

Using the updated authorFacts table to get the number of articles published per year 

```{r}
yeardf <- dbGetQuery(mydb, statement = paste("SELECT DISTINCT PublishYear Year, 
      SUM(TotalPublications) OVER(PARTITION BY PublishYear)as PublicationsPeryear
FROM authorFactsUpd
ORDER BY 1"))
yeardf
```


```{r}
ggplot(yeardf, aes(x=Year, y=PublicationsPeryear, group = 1)) +
  geom_line(color="#69b3a2", size=1) +
  ggtitle("Number of journals published per year") +
geom_text(aes(label=PublicationsPeryear), vjust=-.5)
```

Most journals were published during the year of 2012. A sharp decline in the number of publications is observed from 2012 to 2013.